Prediction of PIK3CA mutation with gene expression

### 2020-05-14 plosOne/plosOne.tex
aurp discussion

Targeted thrapy becomes standard treatment in many cancer patients. Many
targeted therapy requires test for a specific cancer genomic alteration,
to treat the patients. Many direct test for the genomic alteration has
been developed and prooved for their clical utility to discriminate
which patients will be response the targeted therapy.

Machine learning approach has been actively researched to detect the
genomic alterations. Machine learning can build a prediction model from
a large number of predictors such as radiolomic features {[}1{]},
pathology image {[}2{]} or gene expression data {[}3{]}. Because most
direct genomic test are more specific and sensitive than predictive
models, machine learning approach might has limited role in clinical
practice. However the machine learning prediction can be a second best
when the direct test fails.

Prediction RAS pathway activation with gene expression data was done in
previous study. {[}5{]} They trained pancancer The Cancer Genome Atlas
(TCGA) data with a supervised elastic net penalized logistic regression
classifier with stochastic gradient descent. The performance of their
model was 84\% with an area under the receiver operating characteristic
(AUROC) curve and 63\% with an area under the precision recall (AUPR)
curve. The authors suggested their approach can be applied to other
genomic alterations.

PIK3CA is encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase which
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to make phosphatidylinositol 3,4,5-triphosphate
(PIP\textsubscript{3}). Phosphatase and tensin homolog (PTEN) changes
PIP\textsubscript{2} to PIP\textsubscript{3} in contrast PI3K.
PIP\textsubscript{3} is a second mesenger to activate protein kinase B
(AKT) which is a serine/threonine-specific protein kinase. AKT inhibits
apoptosis and promote cell proliferation. {[}6{]}

Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor positive HER2 negative subtype. {[}7{]}
The PIK3CA mutation is second most driver mutation after TP53. The
PIK3CA mutation is most frequantly founded in endometrial carcinoma
(45\%), and followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma and endocervical adenocarcinoma (20\%) and colon
adenocarcinoma (16\%).

We apply a supervised elastic net penalized logistic regression model in
prediction PIK3CA mutation. The purpose of this study is to know this
prediction model approach can be applied not only RAS pathway activation
but also PIK3CA mutation across many cancer types.

@@

We used TCGA pancancer dataset. TCGA is a cancer genomic consotium that
archives data of exom sequencing, gene expression, DNA methylation,
protein expression and clicial data of more than 10000 cancer samples
across 33 common cancer types. The gene expression TCGA pancancer
dataset is batch-corrected with normalization. TCGA dataset is
publically available. PIK3CA mutation data was get using cgdsr
rpackage.{[}8{]} Gene expression data was get from GDAC firehose using
RTCGAToolbox R package. {[}9{]}

10845 cases were available both PIK3CA mutation and mRNA expression
data. 5128 out of 20502 genes were included in the modeling process
after filtering with median absolute deviation as described at modeling
process method. 33 cancer type dummy variables were included in
predictor variables.

The target variable was PIK3CA mutation status. The status of PIK3CA was
considered as positive when the case has following PIK3CA variants which
is the target variables of the therascreen PIK3CA RGQ PCR Kit; C420R,
E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R, H1047Y.
The therascreen PIK3CA RGQ PCR Kit was approved as a companion diagnosis
to treat with PIK3CA inhibitor by U.S. Food and Drug adminidtration. We
splited the three quarters of dataset for the trainset and one quarter
for testset.

@@

Model performance was evealuated with the area under receiver operating
characteristic (ROC) and area under the ROC (AUROC) and precision recall
(PR) and area under curve (AUPR). The prevalence of PIK3CA mutation is
low. The low prevalence of PIK3CA results in imbalaced dataset which
makes the prediction difficult. The AUROC is optimistic in terms of
performance. The AUPR is more informative than AUROC on imbalaced
datasets. {[}12{]} The modeling process was done with tidymodels
rpackage. {[}13{]}

@@

\hypertarget{prevalance-rate-rate-of-pik3ca-mutation}{%
\subsection{prevalance rate rate of PIK3CA
mutation}\label{prevalance-rate-rate-of-pik3ca-mutation}}

Prevalance rate of PIK3CA was 0.11 in all cases. The PIK3CA prevalance
rate of each cancer type was vary. The median prevalance rate of PIK3CA
of each cancer types was 0.03 (range 0-0.33) (Figure 1).

@@

In 10-fold cross-validation, the model with \(\lambda\) = 0.01 and
\(\alpha\) = 1.0 (Ridge regression) showed best performance in terms of
AUROC. The final model was trained with the selected hyperparameters
with all trainset.

@@

Because tht prevalence of PIK3CA mutation is vary across the cancer
type, the performance of each cancer type was investigated. The AUROC
and AUPR were positively correlated between train set and test set in
cancer type subanalysis. (Figure 1B) The AUPR was high in cancer type
with high PIK3CA mutation rate such as colon, brest, Uterus cancer
types. The AUROC did not correlated with PIK3CA mutation rate of each
cancer types. (Figure 1C)

@@

Figure 2 shows top 30 important predictors. The coefficient is the
parameter of the predictor which represent the effect of the predictor
on prediction. IGF1R mRNA expression was the strongest negative
predictor and PTEN was the strongest positive predictor. Both genes and
PIK3CA are key players in tyrosin kanase pathway. The cancer type was
important predictors. Some cancer types including uterine carcinosarcoma
(UCS), bladder urothelial carcinoma (BLCA), pancreatic adenocarcinoma
(PAAD), lymphoid neoplasm diffuse large B-cell lymphoma (DLBC) are
strongest predictors.

@@

\begin{itemize}
\tightlist
\item
  Main message answers the question and main supporting evidence
\end{itemize}

Our model showed good performance to predict PIK3CA mutation of various
cancer types. This result shows that the supervised elastic net
penalized logistic regression model can be applied not only RAS
activation pathway but also PIK3CA mutation. Both RAS activation pathway
and PIK3CA mutation are important and common cancer genomic alterations.
They have significant effect on gene expression in cacer cells. It might
be challenging prediction of genomic alterations which are infrequent or
have weak effect on gene expression.

@@

Prediction modeling from TCGA pancancer dataset has limitations
regarding to data preprocessing. Methods for gancer gene expression has
been developed for research. To stastical analysis, the gene expression
data is processed between-sample normalization to remove batch effect.
If the model has been trained from the between-sample normalization, a
new sample can not be exactly processed like trainset. A model based on
gene expression TCGA pancacner dataset has limitation on preprocessing.
It is nessessory developing preprocessing method which is independent
with dataset to apply the gene expression data to prediction model.

@@

Our PIK3CA prediction model performed better than RAS activation
prediction model of previous study in terms of both AUROC (0.84 vs 0.75)
and AUPR (0.39 vs 0.24) on testset prediction. Our testset is
corresponding to the samples initially filtered from training. The
target variable of our study is more specific than the previous study.
The specific important mutations can effect stronger downsteam gene
expression than the broad events pathway activation. The previous study
might be more difficult prediction problem.

Our model includes cancer type predictor and they are stronger than gene
expression data. The varing prevalence of PIC3CA mutation across cancer
type might reason of the strong cancer type predictor. If the cancer
type was wrong or can not be determined, our model performance can be
poor.

Some significant gene expression predictors were closely related with
PTEN and the PI3K pathway. PTEN and IGFR1R are the strongest gene
expression predictor which have negative and positive predictive power.
IGF1R is a tyrosine kinase receptor which activates PI3K. {[}14{]}
Insulin receptor substrate-2 (IRS2) is the adaptor protein of IGF1R.
{[}15{]} PTEN is an important regulator of PIP\textsubscript{3} by
dephosphorylating PIP\textsubscript{3} in constrast PI3K.{[}6{]}

Another study of PIK3CA mutation prediction showed good performance
AUROC 0.71 in independent testset. They made gene-expression signature
which is sum of the average of the logarithmic gene expression.
{[}4,16{]}

Another study predicted copy numbear alterations with gene expression
using multinomial logistic regression model with least absolute
shrinkage and selection operator (LASSO). The prediction of 1p/19q codel
was very good with an AUROC of 0.997. The gene level prediction was good
with an AUROC 0.75. {[}17{]}

A Hidden Markov Model Approach for Prediction of Genomic Alterations
from Gene Expression Profiling.{[}18{]}

{[}19{]}

\begin{itemize}
\item
  Conclusions comments on possible biological or clinical implications
  and suggestions for further research.
\item
  Proof of concept study\\
\item
  Reproducibility of gene expression prediction model
\end{itemize}

@@

\begin{itemize}
\item
  Figure 1. prevalance rate rate of PIK3CA across cancer types The
  abbreviations of cancer types are explained in S1 appendix.
\item
  Figure 2. Summary of modeling results
\end{itemize}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Left: receiver operating characteristic (ROC) curve right: precision
  recall (PR) curve of trainset and testset. The horizontal green line
  is the PIK3CA mutation rate (0.11) (B) Correlation between trainset
  and testset of area under receiver operating characteristic curve
  (AUROC) and area under precision recall curve (AUPR) among cancer
  type. The abbreviations are explained in S1 appendix. (C) Correlation
  between the PIK3CA mutation rate of area under receiver operating
  characteristic curve (AUROC) and area under precision recall curve
  (AUPR).
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Top 30 high coefficients of mRNA. (B) Coefficients of cancer types.
  The abbreviations of cancer types are explained in S1 appendix.
\end{enumerate}

>>>>

Targeted therapy becomes a standard treatment in many cancer patients.
Much targeted therapy requires a test for a specific cancer genomic
alteration, to treat the patients. Many direct tests for the genomic
alteration have been developed and prooved for their clinical utility to
find which patients will be responding to the targeted therapy.

The machine learning approach has been actively researched to detect
genomic alterations. Machine learning can build a prediction model from
a large number of predictors such as radiolomic features {[}1{]},
pathology image {[}2{]} or gene expression data {[}3{]}. Because the
most direct genomic test is more specific and sensitive than predictive
models, the machine learning approach might have a limited role in
clinical practice. However, the machine learning prediction can be a
second-best when the direct test fails.

Prediction RAS pathway activation with gene expression data was done in
the previous study. {[}5{]} They trained pan-cancer The Cancer Genome
Atlas (TCGA) data with a supervised elastic net penalized logistic
regression classifier with stochastic gradient descent. The performance
of their model was 84\% with an area under the receiver operating
characteristic (AUROC) curve and 63\% with an area under the
precision-recall (AUPR) curve. The authors suggested their approach can
be applied to other genomic alterations.

PIK3CA encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase that
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to make phosphatidylinositol 3,4,5-triphosphate
(PIP\textsubscript{3}). Phosphatase and tensin homolog (PTEN) changes
PIP\textsubscript{2} to PIP\textsubscript{3} in contrast, PI3K.
PIP\textsubscript{3} is a second messenger to activate protein kinase B
(AKT) which is a serine/threonine-specific protein kinase. AKT inhibits
apoptosis and promotes cell proliferation. {[}6{]}

Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor-positive HER2 negative subtype. {[}7{]}
The PIK3CA mutation is the second most driver mutation after TP53. The
PIK3CA mutation is most frequently founded in endometrial carcinoma
(45\%), and followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma, and endocervical adenocarcinoma (20\%) and
colon adenocarcinoma (16\%).

We apply a supervised elastic net penalized logistic regression model in
prediction PIK3CA mutation. The purpose of this study is to know this
prediction model approach can be applied not only to RAS pathway
activation but also PIK3CA mutation across many cancer types.

@@

We used the TCGA pan-cancer dataset. TCGA is a cancer genomic consortium
that archives data of exome sequencing, gene expression, DNA
methylation, protein expression, and clinical data of more than 10000
cancer samples across 33 common cancer types. The gene expression TCGA
pan-cancer dataset is batch-corrected with normalization. The TCGA
dataset is publically available. PIK3CA mutation data was got using
cgdsr rpackage.{[}8{]} Gene expression data was got from the GDAC
firehose using the RTCGAToolbox R package. {[}9{]}

10845 cases were available for both PIK3CA mutation and mRNA expression
data. 5128 out of 20502 genes were included in the modeling process
after filtering with median absolute deviation as described at the
modeling process method. 33 cancer type dummy variables were included in
predictor variables.

The target variable was the PIK3CA mutation status. The status of PIK3CA
was considered as positive when the case has following PIK3CA variants
which are the target variables of the therascreen PIK3CA RGQ PCR Kit;
C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R,
H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved as a companion
diagnostics to treat with PIK3CA inhibitor by the U.S. Food and Drug
Administration. We split the three-quarters of the dataset for the
trainset and one quarter for the test set.

@@

Model performance was evaluated with the area under the receiver
operating characteristic (ROC) and area under the ROC (AUROC) and
precision-recall (PR) and area under the curve (AUPR). The prevalence of
PIK3CA mutation is low. The low prevalence of PIK3CA results in an
imbalanced dataset which makes the prediction difficult. The AUROC is
optimistic in terms of performance. The AUPR is more informative than
AUROC on imbalanced datasets. {[}12{]} The modeling process was done
with tidymodels rpackage. {[}13{]}

@@

\hypertarget{prevalence-rate-of-pik3ca-mutation}{%
\subsection{prevalence rate of PIK3CA
mutation}\label{prevalence-rate-of-pik3ca-mutation}}

The prevalence rate of PIK3CA was 0.11 in all cases. The PIK3CA
prevalence rate of each cancer type varied. The median prevalence rate
of PIK3CA of each cancer type was 0.03 (range 0-0.33) (Figure 1).

@@

In 10-fold cross-validation, the model with \(\lambda\) = 0.01 and
\(\alpha\) = 1.0 (Ridge regression) showed best performance in terms of
AUROC. The final model was trained with the selected hyperparameters
with the entire trainset.

@@

Because the prevalence of PIK3CA mutation varies across the cancer type,
the performance of each cancer type was investigated. The AUROC and AUPR
were positively correlated between the train set and test set in cancer
type subanalysis. (Figure 1B) The AUPR was high in cancer type with high
PIK3CA mutation rate such as colon, breast, Uterus cancer types. The
AUROC did not correlate with the PIK3CA mutation rate of each cancer
type. (Figure 1C)

@@

Figure 2 shows the top 30 important predictors. The coefficient is the
parameter of the predictor which represents the effect of the predictor
on prediction. IGF1R mRNA expression was the strongest negative
predictor and PTEN was the strongest positive predictor. Both genes and
PIK3CA are key players in the tyrosine kinase pathway. The cancer type
was an important predictor. Some cancer types including uterine
carcinosarcoma (UCS), bladder urothelial carcinoma (BLCA), pancreatic
adenocarcinoma (PAAD), lymphoid neoplasm diffuse large B-cell lymphoma
(DLBC) are the strongest predictors.

@@

\begin{itemize}
\tightlist
\item
  The main message answers the question and main supporting evidence
\end{itemize}

Our model showed good performance to predict the PIK3CA mutation of
various cancer types. This result shows that the supervised elastic net
penalized logistic regression model can be applied not only to the RAS
activation pathway but also PIK3CA mutation. Both the RAS activation
pathway and PIK3CA mutation are important and common cancer genomic
alterations. They have a significant effect on gene expression in cancer
cells. It might be a challenging prediction of genomic alterations that
are infrequent or have a weak effect on gene expression.

@@

Prediction modeling from the TCGA pan-cancer dataset has limitations
regarding data preprocessing. Methods for cancer gene expression have
been developed for research. To statistical analysis, the gene
expression data is processed between-sample normalization to remove the
batch effect. If the model has been trained from the between-sample
normalization, a new sample can not be exactly processed like trainset.
A model based on gene expression TCGA pan-cancer dataset has a
limitation on preprocessing. It is necessary for developing a
preprocessing method that is independent with a dataset to apply the
gene expression data to the prediction model.

@@

Our PIK3CA prediction model performed better than the RAS activation
prediction model of the previous study in terms of both AUROC (0.84 vs
0.75) and AUPR (0.39 vs 0.24) on test set prediction. Our test set is
corresponding to the samples initially filtered from training. The
target variable of our study is more specific than the previous study.
The specific important mutations can affect stronger downstream gene
expression than the broad events pathway activation. The previous study
might be a more difficult prediction problem.

Our model includes cancer type predictor and they are stronger than gene
expression data. The varying prevalence of PIC3CA mutation across cancer
type might reason for the strong cancer type predictor. If the cancer
type was wrong or can not be determined, our model performance can be
poor.

Some significant gene expression predictors were closely related to PTEN
and the PI3K pathway. PTEN and IGFR1R are the strongest gene expression
predictor which have negative and positive predictive power. IGF1R is a
tyrosine kinase receptor that activates PI3K. {[}14{]} Insulin receptor
substrate-2 (IRS2) is the adaptor protein of IGF1R. {[}15{]} PTEN is an
important regulator of PIP\textsubscript{3} by dephosphorylating
PIP\textsubscript{3} in contrast PI3K.{[}6{]}

There is some study trying to predict genomic alterations from gene
expression data. A study performed PIK3CA mutation prediction by a
gene-expression signature which is a sum of the average of the
logarithmic gene expression. The model showed good performance AUROC
0.71 in an independent test set. {[}4,16{]} Another study predicted copy
number alterations with gene expression using a multinomial logistic
regression model with least absolute shrinkage and selection operator
(LASSO). The prediction of the 1p/19q codel was very good with an AUROC
of 0.997. The gene-level prediction was good with an AUROC 0.75.
{[}17{]} A Hidden Markov Model Approach for Prediction of Genomic
Alterations from Gene Expression Profiling.{[}18{]} A logistic
regression model was used for MYCN gene amplification in neuroblastoma.
{[}19{]}

\begin{itemize}
\tightlist
\item
  Conclusions comments on possible biological or clinical implications
  and suggestions for further research.
\end{itemize}

Our study suggests that the prediction of genomic alteration with gene
expression data is possible with good performance. However, improvement
of performance is required for applying to clinical test and the
standardization of generation processing of gene expression data is also
needed.

@@

\begin{itemize}
\item
  Figure 1. prevalence rate of PIK3CA across cancer types The
  abbreviations of cancer types are explained in the S1 Appendix.
\item
  Figure 2. Summary of modeling results
\end{itemize}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Left: receiver operating characteristic (ROC) curve right:
  precision-recall (PR) curve of trainset and test set. The horizontal
  green line is the PIK3CA mutation rate (0.11) (B) Correlation between
  trainset and test set of the area under the receiver operating
  characteristic curve (AUROC) and area under the precision-recall curve
  (AUPR) among cancer type. The abbreviations are explained in the S1
  Appendix. (C) Correlation between the PIK3CA mutation rate of area
  under the receiver operating characteristic curve (AUROC) and area
  under the precision-recall curve (AUPR).
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Top 30 high coefficients of mRNA. (B) Coefficients of cancer types.
  The abbreviations of cancer types are explained in the S1 Appendix.
\end{enumerate}
